D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-04-05 03labelled.tex
fix typos
.

For given context $\Gamma$ and any valuation $v$,
we interpret 
$$\mbox{$\sem{\cD(s,s')}_v \triangleq \mathcal{L}(\bigotimes_{j}\cH_{\sem{\sigma_{r'_j}}_v}, \bigotimes_{i}\cH_{\sem{\sigma_{r_i}}_v})$}$$ where $\mathcal{L}$ denotes the set of linear maps, 
$s = \{r_1,\cdots,r_n\}$ and $s' = \{r'_1,\cdots,r'_m\}$ ($r_i$ and $r'_j$ are sorted).
% we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty. \label{context: sigma s}
% For any valuation $v$, we further write $\cH_{\sem{s}_v} \triangleq \bigotimes_{i}\cH_{\sem{\sigma_i}_v}$ which equals to $\cH_{\sem{\sigma_s}_v}$ for non-empty $s$.
% We interpret $\sem{\cD(s,s')} \triangleq \mathcal{L}(\cH_{\sem{s'}_v}, \cH_{\sem{s}_v})$ where $\mathcal{L}$ denotes the set of linear maps.
% for any $s\subseteq \cR$, where $\Gamma\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. 
We interpret labelled Dirac notations inductively as (assume $\Gamma\vdash D_i : \cD(s_i,s'_i)$) :
\begin{itemize}
  \item $\sem{|i\>_r}_v = |\sem{i}_v\>$;\quad 
        $\sem{\<i|_r}_v = \<\sem{i}_v|$;\quad  
        $\sem{K_R}_v = \sem{\Swap_R}_v \cdot \sem{K}_v$; \\
        $\sem{B_R}_v = \sem{B}_v\cdot \sem{\Swap_R}_v$;\quad
        $\sem{O_{R_1,R_2}}_v = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v$;
  \item $\sem{D_1\otimes \cdots \otimes D_n}_v = \sem{\Swap_{s_1,\cdot,s_n}}_v\cdot (\sem{D_1}_v\otimes \cdots \otimes \sem{D_1}_v)\cdot \sem{\Swap_{s'_1,\cdot,s'_n}}_v$;
  \item $\sem{D_1\cdot D_2}_v = \sem{\cl(D_1, s_2\backslash s'_1)}_v\cdot \sem{\cl(D_2, s'_1\backslash s_2)}_v$. Note that $s_2\backslash s'_1$ and $s'_1\backslash s_2$ are the minimal extension that make it interpretable. E.g., to interpret ${}_p\<i|\cdot |j\>_{p,q}$, we at least need to extend ${}_p\<i|$ to ${}_p\<i|\otimes I_{q}$.
\end{itemize}
It can be shown that Lemma \ref{lemma:sound type system} also holds for labelled terms, i.e., $\sem{D}_v \in \sem{\cD(s,s')}_v$ given $\Gamma\vdash D : \cD(s,s')$. Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2}_v = \sem{D_2\otimes D_1}_v$ and $\sem{D_1\otimes (D_2\otimes D_3)}_v = \sem{(D_1\otimes D_2)\otimes D_3}_v$, which ensures the soundness of treating labelled tensor as an AC symbol.

@@

\paragraph*{2. Rewriting to the normal form.}
We add three types of rules for dealing with operators on labelled terms, 1) recursively applying them to subterms, 2) pushing big operators out and 3) eliminating generalized composition and bra-ket pairs.
%Other symbols on labelled Dirac notation are also 
Take the rule for conjugate $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$ as an example of 1).
For 2), we use distributivity rules for scaling, labelled tensor and generalized composition. For example, rule
\[
% \textrm{(R-SUM-PUSHD0)}
X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n)
\]
will lift summation to the outside.
Extra rules for 3) are established including:
% The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
\begin{align*}
    % & \textrm{(R-SUM-PUSHD0)}
    % && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
    % %
    % & \textrm{(R-SUM-PUSHD1)}
    % && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
    %
    & \textrm{(R-L-SORT0)}
    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
    %
    & \textrm{(R-L-SORT1)}
    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
    %
    & \textrm{(R-L-SORT2)}
    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)
    %
\end{align*}
% These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
Assuming no variables of $\DType(s_1, s_2)$, 
repeating the application of the above rules yield the normal form of both sides of the equality--the addition of big operators: 
\begin{equation}
  \label{eqn:Labelled normal}
  \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
  + \cdots +
  \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
\end{equation}
where each sum body is a tensor of labelled basis with scalar coefficients in plain Dirac notation.
\paragraph*{3. Ordering and elimination of quantum variables.} 
We further sort tensors of labelled basis in every sum body of Eqn. (\ref{eqn:Labelled normal}) by 1. ket first and 2. the default order of variables. This yield the same shape of every subterm on both sides of the equality, e.g., 
\begin{equation}
  \label{eqn:Labelled normal1}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i}_p \otimes \cdots \otimes \ket{j}_q) \otimes (\bra{i'}_{p'}\otimes\cdots\otimes \bra{j'}_{q'}))
\end{equation}
and thus it is equivalent to decide the equivalence of additions of subterms of:
\begin{equation}
  \label{eqn:Labelled normal2}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i} \otimes \cdots \otimes \ket{j}) \cdot (\bra{i'}\otimes\cdots\otimes \bra{j'}))
\end{equation}
which do not involve any labels.

>>>>

For given context $\Gamma$ and any valuation $v$,
we interpret 
$$\mbox{$\sem{\cD(s,s')}_v \triangleq \mathcal{L}(\bigotimes_{j}\cH_{\sem{\sigma_{r'_j}}_v}, \bigotimes_{i}\cH_{\sem{\sigma_{r_i}}_v})$}$$ where $\mathcal{L}$ denotes the set of linear maps, 
$s = \{r_1,\cdots,r_n\}$ and $s' = \{r'_1,\cdots,r'_m\}$ ($r_i$ and $r'_j$ are sorted).
% we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty. \label{context: sigma s}
% For any valuation $v$, we further write $\cH_{\sem{s}_v} \triangleq \bigotimes_{i}\cH_{\sem{\sigma_i}_v}$ which equals to $\cH_{\sem{\sigma_s}_v}$ for non-empty $s$.
% We interpret $\sem{\cD(s,s')} \triangleq \mathcal{L}(\cH_{\sem{s'}_v}, \cH_{\sem{s}_v})$ where $\mathcal{L}$ denotes the set of linear maps.
% for any $s\subseteq \cR$, where $\Gamma\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. 
We interpret labelled Dirac notations inductively as (assume $\Gamma\vdash D_i : \cD(s_i,s'_i)$) :
\begin{itemize}
  \item $\sem{|i\>_r}_v = |\sem{i}_v\>$;\quad 
        $\sem{\<i|_r}_v = \<\sem{i}_v|$;\quad  
        $\sem{K_R}_v = \sem{\Swap_R}_v \cdot \sem{K}_v$; \\
        $\sem{B_R}_v = \sem{B}_v\cdot \sem{\Swap_R}_v^\dagger$;\quad
        $\sem{O_{R_1,R_2}}_v = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v^\dagger$;
  \item $\sem{D_1\otimes \cdots \otimes D_n}_v = \sem{\Swap_{s_1,\cdot,s_n}}_v\cdot (\sem{D_1}_v\otimes \cdots \otimes \sem{D_1}_v)\cdot \sem{\Swap_{s'_1,\cdot,s'_n}}_v^\dagger$;
  \item $\sem{D_1\cdot D_2}_v = \sem{\cl(D_1, s_2\backslash s'_1)}_v\cdot \sem{\cl(D_2, s'_1\backslash s_2)}_v$. Note that $s_2\backslash s'_1$ and $s'_1\backslash s_2$ are the minimal extension that make it interpretable. E.g., to interpret ${}_p\<i|\cdot |j\>_{p,q}$, we at least need to extend ${}_p\<i|$ to ${}_p\<i|\otimes I_{q}$.
\end{itemize}
It can be shown that Lemma \ref{lemma:sound type system} also holds for labelled terms, i.e., $\sem{D}_v \in \sem{\cD(s,s')}_v$ given $\Gamma\vdash D : \cD(s,s')$. Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2}_v = \sem{D_2\otimes D_1}_v$ and $\sem{D_1\otimes (D_2\otimes D_3)}_v = \sem{(D_1\otimes D_2)\otimes D_3}_v$, which ensures the soundness of treating labelled tensor as an AC symbol.

@@

\paragraph*{2. Rewriting to normal form.}
We add three types of rules for dealing with operators on labelled terms, 1) recursively applying them to subterms, 2) pushing big operators out and 3) eliminating generalized composition and bra-ket pairs.
%Other symbols on labelled Dirac notation are also 
Take the rule for conjugate $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$ as an example of 1).
For 2), we use distributivity rules for scaling, labelled tensor and generalized composition. For example, rule
\[
% \textrm{(R-SUM-PUSHD0)}
X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n)
\]
will lift summation to the outside.
Extra rules for 3) are established including:
% The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
\begin{align*}
    % & \textrm{(R-SUM-PUSHD0)}
    % && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
    % %
    % & \textrm{(R-SUM-PUSHD1)}
    % && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
    %
    & \textrm{(R-L-SORT0)}
    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
    %
    & \textrm{(R-L-SORT1)}
    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
    %
    & \textrm{(R-L-SORT2)}
    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)
    %
\end{align*}
% These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
Assuming no variables of $\DType(s_1, s_2)$, 
repeating the application of the above rules yield the normal form of both sides of the equality--the addition of big operators: 
\begin{equation}
  \label{eqn:Labelled normal}
  \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
  + \cdots +
  \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
\end{equation}
where each sum body is a tensor of labelled basis with scalar coefficients in plain Dirac notation.
\paragraph*{3. Ordering and elimination of quantum variables.} 
We further sort tensors of labelled basis in every sum body of Eqn. (\ref{eqn:Labelled normal}) by 1. ket first and 2. the default order of variables. This yield the same shape of every subterm on both sides of the equality, e.g., 
\begin{equation}
  \label{eqn:Labelled normal1}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i}_p \otimes \cdots \otimes \ket{j}_q) \otimes (\bra{i'}_{p'}\otimes\cdots\otimes \bra{j'}_{q'}))
\end{equation}
and thus it is equivalent to decide the equivalence of additions of subterms of:
\begin{equation}
  \label{eqn:Labelled normal2}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i} \otimes \cdots \otimes \ket{j}) \cdot (\bra{i'}\otimes\cdots\otimes \bra{j'}))
\end{equation}
which do not involve any labels.

### 2025-04-05 05tool_eval.tex
.

\section{Implementation and Case Study}
The improvement and extensions discussed above culminate in our implementation, called \textbf{D-Hammer}, a solver written in \CC. 
It features a parser built using ANTLR4, and scalar reasoning is powered by the Mathematica Engine. Users can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output. 
D-Hammer can be run interactively from the command line or integrated into other \CC\ projects as a library.
This tool has been tested on benchmarks from previous work and new examples for labelled Dirac notation, showing significant improvements in expressivity and efficiency. 
The tool, along with example encodings, will be made open-source and publicly available.

@@

\subsubsection{Textbook (QCQI)}
The 18 examples of QCQI textbook equations from DiracDec are all encoded and solved. The running time are short and close because these examples are simple and direct.

@@

\subsubsection{CoqQ}
In terms of expressibility, D-Hammer encodes 158 examples for the CoqQ part, which is 4 fewer than DiracDec. This discrepancy arises because DiracDec supports projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, where $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$. However, we found that this feature is rarely used and removed the support in D-Hammer. Regarding decidability, we improved the rewriting rules for sums in D-Hammer, and proved several more examples failed by DiracDec. 
\Cref{fig: CoqQ plot} shows a direct comparison of their time efficiency.
For small examples, D-Hammer is slower due to the marginal overhead. 
However, as running time increases, D-Hammer accelerates by an average factor of 2 to 40 times.
The relationship between efficiency improvement and example characteristics further highlights the necessity of our algorithm. 
Notably, examples with great efficiency improvement, e.g. COQQ-129 and COQQ-148 shown in~\Cref{fig: CoqQ plot}, tend to use deeply nested sums, where our algorithm has a more pronounced effect.

>>>>

\section{Implementation and Case Study}
We present \textbf{D-Hammer}, an open-source and publicly
available\footnote{{\textcolor{red}{URL}}} implementaion of our
approach.  \textbf{D-Hammer} is an equational prover for Labelled
Dirac notation written in \CC.  It features a parser built using
ANTLR4, and scalar reasoning is powered by the Mathematica
Engine. Users can use commands to make definitions and assumptions in
the maintained context, conduct the normalization and equivalence
checking, and obtain the rewriting trace output.  D-Hammer can be run
interactively from the command line or integrated into other
\CC\ projects as a library.

@@

The results show significant performance improvements. The enhanced
efficiency primarily results from algorithmic improvements rather than
the shift to a C++ implementation. This is supported by the fact that
efficiency gains are more significant for larger examples, as can be
observed in \Cref{fig: CoqQ plot}.

\subsubsection{Textbook (QCQI)}
As a warm-up, we consider 18 examples from Nielsen and Chuang's
classic texbook~\cite{nielsen2010quantum}. All examples can be encoded
in DiracDec and D-Hammer and are solved very very efficiently.

@@

\subsubsection{CoqQ}
As a more substantial example, we consider the examples from
CoqQ~\cite{Zhou2023}, an extensive formalization of quantum
information theory and quantum programming languages in the
Coq proof assistant.

CoqQ has been used as the main benchmark for evaluating DiracDec.
Specifically, \cite{diracdec} isolates 162 statements in CoqQ that
are in scope of the DiracDec language.

We have ported 158 out of 162 examples to D-Hammer. The remaining 4
examples uses projectors $\texttt{fst}$ and $\texttt{snd}$ on basis
pairs, where $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) =
t$. However, we found that this feature is rarely used and removed the
support in D-Hammer. Note that the omission of the projection rules
has a limited influence on the performance evaluation, because the
efficiency of the 158 examples is not affected by projection rules.

D-Hammer verifies the whole 158 examples in less than 10 seconds,
whereas DiracDec verifies 156/162 examples in more than 45 seconds.
\Cref{fig: CoqQ plot} shows a direct comparison of the efficiency of
the two tools. We observe that D-Hammer is slower than DiracDec on
small examples, due to marginal overhead, but becomes faster by an
average factor of 2 to 40 times as the running time of examples
increases. This non-linear growth suggests that efficiency gains
result from algorithmic improvements rather than the shift to a C++
implementation. Furthermore, examples with great improvements,
e.g. COQQ-129 and COQQ-148 shown in~\Cref{fig: CoqQ plot}, tend to use
deeply nested sums, for which our algorithm is more efficient.

### 2025-04-05 06related_conclusion.tex
.

\section{Related Work}
Categorical quantum mechanics (CQM)~\cite{DBLP:conf/lics/AbramskyC04}
is a category-theoretic foundation for quantum mechanics. CQM is the
basis of the ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08}, a popular
graphical calculus for quantum states~\cite{vandewetering2020zx}. A
key aspect of the ZX calculus and related graphical calculi is that
they can be equipped with powerful rewriting systems. These systems
support a form of graph rewriting, i.e.\, they exploit the graph
representation of quantum states. There exists several completeness
results for ZX-like rewriting systems, as well as some practical
tools, such as PyZX~\cite{kissinger2019pyzx}. Similarly, there exist
completeness results and tools for the sum-over-paths representation
of quantum states~\cite{amy2018towards,amy2023complete}.

Automata-based approaches has also been successfully applied for reasoning about quantum circuit~\cite{AutoQ_pldi_2023}. Instead of representing a set of  $n$-qubit states in exponential size, \cite{AutoQ_pldi_2023} encode it as (finite) tree automata with much less (often $\mathit{poly}(n)$) transitions.  They further established transformers to implement the semantics of quantum gates over this representation, providing better scalability compared to other approaches. It is developed as a practical tool \textsc{autoq} with the capabilities for symbolic reasoning~\cite{AutoQ2023}. More recently, \cite{AutoQ_popl2025} further explore level-synchronized tree automata to encoding state, a promising model for parameterized verification.

@@

The most obvious difference is that D-Hammer targets labelled Dirac
notation, whereas DiracDec targets plain Dirac notation. As already
mentioned, the use of labelled Dirac notation is essential in many
applications; in particular, labelled Dirac notation can simplify
notation and proofs, and is in general better suited for writing and
reasoning about complex, many-body, quantum states. However, there are
several other important differences. First, our language leverages the
power of higher-order dependent type theory. In particular, we
leverage higher-order functions to provide a compact and expressive
representation of big operators. Second, our language adopts AC
symbols with indefinite arities, which leads to compact representation
of terms, and eases AC reasoning. Third, the relation with Mathematica
is fundamentally different. DiracDec is implemented in Mathematica; as
a consequence, its behavior, and in particular, its typing rules are
constrained by the lack of typing in Mathematica. In contrast, D-Hammer
is designed as a separate tool; as a consequence, it benefits from an
improved representation of terms (e.g.\, AC operators with indefinite
arity), a more expressive type system (e.g.\, dependent types), and a
more efficient rewriting engine. D-Hammer still relies on Mathematica
to reason about functions that are not natively supported by rewriting
rules, but these interactions are constrained and do not have negative
effects on the overall efficiency of the system. This is reflected in
our experimental comparison of D-Hammer and DiracDec, which shows how
the former outperforms the latter.

>>>>

\section{Related Work}

@@

The most obvious difference is that D-Hammer targets labelled Dirac
notation, whereas DiracDec targets plain Dirac notation. As already
mentioned, the use of labelled Dirac notation is essential in many
applications; in particular, labelled Dirac notation can simplify
notation and proofs, and is in general better suited for writing and
reasoning about complex, many-body, quantum states. However, there are
several other important differences. First, our language leverages
higher-order functions to provide a compact and expressive
representation of big operators. Second, our language adopts AC
symbols with indefinite arities, which leads to compact representation
of terms, and eases AC reasoning. Third, the relation with Mathematica
is fundamentally different. DiracDec is implemented in Mathematica; as
a consequence, its behavior, and in particular, its typing rules are
constrained by the lack of typing in Mathematica. In contrast,
D-Hammer is designed as a separate tool; as a consequence, it benefits
from an improved representation of terms (e.g.\, AC operators with
indefinite arity), a more expressive type system (e.g.\, dependent
types), and a more efficient rewriting engine. D-Hammer still relies
on Mathematica to reason about functions that are not natively
supported by rewriting rules, but these interactions are constrained
and do not have negative effects on the overall efficiency of the
system. This is reflected in our experimental comparison of D-Hammer
and DiracDec, which shows how the former outperforms the latter.

\paragraph*{Comparison with ZX Calculus}
The ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08,vandewetering2020zx}
is a popular graphical calculus for reasoning about quantum states.
The foundations of the ZX calculus are grounded in categorical quantum
mechanics (CQM)~\cite{DBLP:conf/lics/AbramskyC04}, which provides an
category-theoretic foundation for quantum mechanics.  A key aspect of
the ZX calculus and related graphical calculi is that they can be
equipped with powerful rewriting systems. These systems support a form
of graph rewriting, i.e.\, they exploit the graph representation of
quantum states. Two notable properties of these systems are:
\begin{itemize}
\item completeness:
\item efficiency:
\end{itemize}
Both Labelled Dirac Notation (LDN) and the ZX calculus share the use
of rewriting techniques. However, it is difficult to compare the ZX
calculus and LDN. First, they have different scopes: the ZX calculus
mainly focuses on quantum circuits, whereas LDN aims to provide
general support for quantum physics. This difference in scope is
reflected in the underlying languages: for instance, many works in ZX
calculus consider concrete sets of quantum gates, whereas LDN
emphasizes the use of big operators, and does not attempt to provide
specialized support for a particular set of quantum gates. As a
result, many of the examples handled by D-Hammer cannot be formalized
in existing ZX-calculus tools. On the contrary, D-Hammer is not
competitive on the examples that can be verified by ZX-calculus.

\gb{give a few examples in terms of size and time}

\paragraph*{Comparison with path-over sums approach}
Similarly, there exist
completeness results and tools for the sum-over-paths representation
of quantum states~\cite{amy2018towards,amy2023complete}.

\paragraph*{Comparison with equivalence checking}

Automata-based approaches has also been successfully applied for
reasoning about quantum circuits~\cite{AutoQ_pldi_2023}. Instead of
representing a set of $n$-qubit states in exponential size,
\cite{AutoQ_pldi_2023} encode it as (finite) tree automata with much
less (often $\mathit{poly}(n)$) transitions.  They further established
transformers to implement the semantics of quantum gates over this
representation, providing better scalability compared to other
approaches. It is developed as a practical tool \textsc{autoq} with
the capabilities for symbolic reasoning~\cite{AutoQ2023}. More
recently, \cite{AutoQ_popl2025} further explore level-synchronized
tree automata to encoding state, a promising model for parameterized
verification.

\paragraph*{Canonical forms in multi-body quantum physics}

\paragraph*{Comparison with egraphs}
Our algorithm is based on term rewriting. However, it is a challenge
to device well-behaved and efficient term rewriting for (labelled)
Dirac notation. An alternative to term rewriting would be to use
equality saturation~\cite{DBLP:journals/pacmpl/WillseyNWFTP21}, a
powerful equational reasoning technique that does not require
existence of normal forms. Equality saturation may be particularly
useful when considering further extensions of labelled Dirac notation.
